---
title: "MTH304"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true 
---
```{r, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r Preamble, eval=TRUE, echo=FALSE, error=FALSE, message=FALSE}
library(gridExtra)
library(MASS)
library(ISLR)
library(car)
library(modelr)
library(gapminder)
library(broom)
library(e1071)
library(summarytools)
library(tidyverse)
library(knitr)
library(png)
library(jpeg)

set.seed(19971222)
```

# Course Description
**Probability and Statistics I:** Topics include elements of Probability Theory. Discrete Probability Distribution. (Hyper-geometric, Binomial, Poisson). Normal Distribution and its applications. Lognormal Distribution, Multivariate Distributions, Covariance and Correlation, Moment Generating Functions, Central limit theorem and applications. A statistics computer package will be used in this course. As copied from the [Ryerson Calendar](https://www.ryerson.ca/calendar/2018-2019/courses/mathematics/MTH/304/).

**Professor** [Dr. You Liang](https://math.ryerson.ca/people/yliang.html)

# Week 1

Coming soon...