---
title: "STA302"
output: 
  html_document:
    toc: TRUE 
    toc_depth: 4
    toc_float: TRUE  
    fig_caption: TRUE 
    df_print: tibble 
---
```{r, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r Preamble, eval=TRUE, echo=FALSE, error=FALSE, message=FALSE}
library(gridExtra)
library(MASS)
library(ISLR)
library(car)
library(modelr)
library(gapminder)
library(broom)
library(e1071)
library(summarytools)
library(tidyverse)
library(knitr)
library(png)
library(jpeg)

set.seed(19971222)
```

# QQ Plots

## A Little Background

In statistics, the first real tool (and often most powerful) for data analysis that becomes available to us is *linear regression*. However, it doesn't come without some catches. To get all of the nice properties proved in class we must ensure that *all* of the assumptions are met. Recall from class the following: 

***Linearity** - dependent variable is a linear combination of the parameters of the model (we refer to them as $\beta$'s) and independent variables. _Aside: this is one of my favourite assumptions as most people often forget to mention this. Linear relationships are not the only possibilities and this will be explored in other courses dedicated to breaking this exact assumption. See STA303_

***Constant Independent Variable** - mathematically we denote this as $\mathbb{E}(e_i) = 0 \text{  } \forall i$ and the underlying meaning of this is that we can treat the recorded independent variable as a known constant

***Independence of errors** - mathematically we write this as $Cor(e_i, e_j) = 0 \text{  } i \neq j$ 

***Constant Variance** - otherwise known as homoskedasticity (love this word, learning how to pronounce it is the first barrier to a statistics degree) and is denoted $Var(e_i) = \sigma^2 < \infty \text{  } \forall i$. 

A nice neat way to package this all up is to write the following (in lecture we called these the Gauss-Markov assumptions).

\begin{equation} 
  \label{eq:1}
  e_i \sim_{iid} N(0, \sigma^2) \text{  } \forall i
\end{equation}

The purpose of this tutorial is to explore one of the ways that we will examine the above via a visual tool we refer to as a Qunatile to Quantile plot. 

## Does this look normal?

As the name suggests we will be working with the quantiles of a some distribution(s). We will develop this idea in the general setting because this works for **any** distribution, then focus specifically on the normal distribution since this is what we care about for this course. Let's get started. 

### The Data

We are recording measurments of some random variable $X$. We would like to check if these observations come from some ideal/suspected distribution $F$. This means that we are *asking* if $X \sim F$. However we are not sure... yet. What we do have is knowledge of what $F$ is since this is what we would like it to be (need to guess something you know) and we have empirical observations of our data. We denote the empirical distribution as $G$.

### If you build it, it might be normal

That is all to say that an equivalent question to the one above is if $F = G$ and what's left is to compare them visually. I'd now like to highlight that $G$ is an empircal cumulative distribution function (ECDF) while $F$ is a cumulative distribution function (CDF) which makes them a bit harder to compare. ECDFs are never smooth since they are based on a finite number of points (as dictated by the data set) while CDFs are beautiful functions that make us happy. Thus, we must find a way to discretize $F$ in order to compare them on the same basis (the beauty of $F$ is that we can do exactly this). The solution is to use the inverse of the distributions, otherwise known as the qunatile function and is denoted $F^{-1}$ and $G^{-1}$ respectively. For our observed data this is easy since we can take the [ordered statistic](https://en.wikipedia.org/wiki/Order_statistic) $x_{(i)}$ as a qunatile and we are done. For $F$ we must do a bit more work (luckily, it's not hard work). Generate a data set (we call this simulation) of equal size as $X$ from $F$ using your  computer and take the corresponding ordered statistics $f_{(i)}$. If we don't exactly match the quantiles of the ECDF we can do some interpolation on the generated data set due to the continuity of $F$ (when data sets are very large this step is likely not needed). 

The last question we must ask ourselves is what it does equality look like geometrically or on the plane. If I was to ask you to draw me a function that is the definition of two quantities being equal, what would it be? The function that should come to mind is $y=x$. By extension we if are to compare the two distribution on a graph we should check if they fall on that line. From this, the interpretation becomes vey natural. If the two distributions in question are indeed the same, the quantiles of each one will fall along the $y=x$ line. Deviations from this line mean that they are not the same. I would like to take this opportunity to emphasize that some deviation is okay as we are dealing with **random** variables afterall. Large, systematic deviations (particularly in the tails) is what would be cause for concern. 

Here is a summary on how we construct the QQ plots.

1. Compute the ECDF of our data set and determine it's quantiles.
2. Compute the quantiles of our theoretical distribution.
3. Plot the theoretical (y axis) vs. the empirical (x axis).
4. Compare them to the $y=x$ line and draw appropriate conclusions. 

### A house not made of straws

When reading this, you will notice the deliberate omission of any particular distribution we are evaluating. All that was required was that $F$ and $G$ are **some** valid probability distribution. This makes QQ plots such a versatile tool because we can test for any distribution we are curious about. In our case, we are intersted in whether or not the errors come from the normal distribution in equation \ref{eq:1} 1. But we can just replace that with your favourite student's t distribution, or F distribution, chi-squared, gamma, or any other you want. You even have the power to blindly compare two ECDF's by letting $F$ be an ECDF. If all you are intereted in is whether or not two datasets follow the same (all be it unkown) distribution this is a solid method to explore that making this a very robust tool.
