---
title: "STA314"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true 
---
```{r, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r Preamble, eval=TRUE, echo=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(gridExtra)
```
Note: Majority of this material is credited to [Prof. Daniel Simpson](http://utstat.toronto.edu/?page_id=11779) and editted by me in cases where I thought further clarification is needed.

## Week 1

This week is centred around learning the basics of R. Check out some of the cheat sheets on the main page to get a better feel for working with this langauge.

## Week 2

### Question1

```{r Question 1, echo=FALSE, eval=FALSE}
setwd("~/Documents/U of T/STA314/Week 2/Week 2 Tutorial")

# load data
load("~/Documents/U of T/STA314/Week 2/Week 2 Tutorial/question1.rdata")
# compute disim using euclidean norm
disimilarity <- dist(dat)

# compute clusters based on different linkage methods 
single <- hclust(d=disimilarity, method = "single")
cmplt <- hclust(d=disimilarity, method = "complete")

# compute kmeans with k=2 and various nstarts
dat$single_linkage <- cutree(tree = single, k=2)
dat$cmplt_linkage <- cutree(tree = cmplt, k = 2)
```

### Question 2

```{r Question 2, eval=TRUE}
### create a function that implements the K-Means algo referenced in class
my_kmeans <- function(data, k, n_starts) {
	
done = FALSE # define this for our 'exit' clause
n = dim(data)[1] #data is a matrix, where each row is one data point

# solve the trivial case in its entirety 
if (k == 1) {
	cluster = rep(1, n) # this vector says which cluster each point is in
	centers = apply(
		X = data,
		MARGIN = 2,
		FUN = mean
		) # use the apply function to calculate the cluster centre
	cost = sum((data - centers[cluster]) ^ 2) # evaluatae cost function
	return(list(cluster = cluster, cost = cost)) # output everything you've calculated
	# in order to evaluate the effectiveness of your model
}

cluster_old = rep(1, n) # "guess" the clustering to give us an inital set up
cost_old = Inf # set our cost to infinity for now 

# now we write a loop in order to get our clusters
for (run in 1:n_starts) {
	cluster = rep(1, n) #this vector says which cluster each point is in
	#uniformly choose initial cluster centers
	centers = data[sample(x = 1:n, 
	                      size = k,
	                      replace = FALSE), ] 
	# sample k random points from our data as the initial clusters
	while (!done) {
		# Do Step 2.1
		d = matrix(nrow = n, ncol = k)
		for (j in 1:k) {
		d[, j] = apply(
			X = data,
			MARGIN = 1,
			FUN = function(d) sum((d - centers[j, ]) ^ 2) 
			# computes cost functions if x_i belongs to each of the centres 
			)
		}
		cluster_new = apply(
			X = d,
			MARGIN = 1,
			FUN = which.min # find the min cost and assign the x_i to that cluster
			)
if (length(unique(cluster_new)) < k)
	stop("Empty cluster!")

# Do Step 2.2
		for (i in 1:k) {
			centers[i, ] = apply(
				X = data[cluster_new == i, ],
				MARGIN = 2,
				FUN = mean) # computes mean of each cluster for each cluster
			}
# Check if the cluster assignements changed. If they have, set done=TRUE
		if (all(cluster == cluster_new)) {
			done = TRUE # if all our clusters were correct then we're done
			}
		cluster = cluster_new # if not then we reassign the values based on the middle points 
	} #end of while not done
	
cost = sum((data - centers[cluster, ]) ^ 2)
if (cost_old < cost) {
	cluster = cluster_old
	cost = cost_old
	}
	
	cost_old = cost
	cluster_old = cluster
}
return(list(cluster = cluster, cost = cost))
}

```

Use this algorithm to make a 4 clustering of the data set in question2.RData. Comment on the clustering.

```{r Q2 continued}

# Load the data
load("~/Documents/U of T/Teaching/STA314/Week 2/Week 2 Tutorial/question2.rdata")

out = my_kmeans(dat_q2,4,2)
dat_q2$cluster = out$cluster
dat_q2 %>% ggplot(aes(x = x,y = y)) +
	geom_point(aes(colour = factor(cluster)))

```

Depending on how many times the kmeans algorithm is run, it sometimes doesn’t find all four distinct clusters.
This is due to the uniform intial sampling and the fact that the bottom left and top right clusters are much
smaller than the other two.

### Question 3

```{r Question 3}
#### PARTIAL SOLUTION ####

# Input the data
d = matrix(c(0, 0.3, 0.4, 0.7,
0.3, 0, 0.5, 0.8,
0.4, 0.5, 0.0, 0.45,
0.7, 0.8, 0.45, 0.0), nrow = 4)

# Set it as distance
d = as.dist(d)
```

```{r Q3 Complete Linkage}
#### PARTIAL SOLUTION ####
# Plot the clusters with complete linkage:
plot(hclust(d,method = "complete"))
```

```{r Q3 Single Linkage}
#### PARTIAL SOLUTION ####
# Plot the clusters with complete linkage:
plot(hclust(d,method = "single"))
```

Comparing the two dendrograms, we see that the two clustering from the complete linkage is \{1,2\}, \{3,4\},
while the two clustering from the single linkage is \{1,2,3\}, \{4\}.

### Question 4 

For part a) there is not enough information. If the two linkages are equal, then they will fuse at the same
hight. Otherwise, the single linkage dendrogram will merge at a lower hight as it only requires one nearby point and not all of the points to be close.

For part b) They’ll merge at the same hight because when you’re just merging single leaves, the linkages all
reduce to the distance and are therefore equal.

## Week 3
This week we covered some of fundamental proofs in linear algebra that have very powerful implications in machine learning. The hardest part of these proofs is the notational aspect of them. If you are unclear about these, please bring them up in next week's tutorial or go to office hours. 
\newline
I focused on Q3 off of the assignment. Here are links to the proofs.
###Q uestion 3

####P art A
**WTS:** $u^Tv = 0 \Rightarrow ||u+v||^2 = ||u||^2 + ||v||^2$ \newline
[Solution](https://math.stackexchange.com/questions/11509/how-to-prove-the-pythagoras-theorem-using-vectors) \newline

#### PART B
**WTS:** $\Phi \in R^{n \times r}$ with orthonormal columns $\Rightarrow ||\Phi v|| = ||v||$ \newline
[Solution](http://www.fen.bilkent.edu.tr/~ercelebi/rotation%20of%20coordinate%20axes.pdf): note that this proof is a bit more complicated than it has to be. The essence is the same, they just break it down component wise (this is an unecessary complication as vector notation makes this proof about 2-3 lines long). \newline

#### Part C
**WTS:**$u,v \in R^n$, what is the largest eigenvalue? \newline
[Solution](https://ccrma.stanford.edu/~dattorro/simplemat.pdf): See part B1. \newline

#### Part D
**WTS:** $A$ is a symmetric matrix, $\max_{x} \frac{x^T A x}{x^Tx}$ corresponds to the largest eigenvalue \newline
[Solution](https://math.stackexchange.com/questions/1705386/showing-that-x-topax-is-maximized-at-max-lambdaa-for-symmetric-a): This proof is almost trivial once you understand spectral decomposition. More on this later. \newline

### Part E
**WTS:** $A$ is symmetric positive definite $\Rightarrow A^TA$ also symmetric positive definite \newline
**Solution:** For this proof I will be using the [Spectral Decomposition Thoerem](http://pages.mtu.edu/~msgocken/ma5630spring2003/lectures/spectral/spectral/node2.html) .
$$A = UDU^T \Rightarrow A^T A = \left( UDU^T \right)^T(U D U^T) = UDU^T UDU^T = UD^2U $$
We get the above since $U^T = U^{-1}$ by definition (take some time to convince yourself of this fact). Furthermore, since D is the diagonal matrix with entries being the eigenvalues, the square of set eigenvalues are necessarily positive which is sufficient to define the positive definite part of the proof. Lastly, by the construction of U the symmetry property follows. This completes the proof. \blacksquare

## Week 4 
This week we focused on Principle Component Analysis (will be referred to as PCA from here on out). PCA is a very useful tool for pulling out the key features from your data. The reason this is important is that, often times, you will be given data sets with a large number of columns corresponding to features. For example, if we were working on predicting heart disease for a given person and were tasked with creating this model, we could be given information such as the person's height, weight, age, hair colour, daily caloric intake, fat consumption, transfat consumption, blood sugar level, cholesterol, etc. Needless to say we probably don't need \textit{every} one of these factors. With this in mind we will leverage our knowledge of eigenpairings in order to reduce the number of factors we will use in our model to ensure we only include the most important ones (the definition of most important is up for debate). 
\newline 
Given that background, we will go through some of the problems from this week's problem set. 
### Question 2 
Please see [Michal's](https://michalmalyska.github.io/STA314.html) for sample code to these labs.

### Question 3 
\textbf{WTS:} The second principle component of a centred feature matrix $X$ should maximize Rayleigh's quotient up to being orthogonal to the fist principle component. 
\newline
\textit{Proof:} Note, this argument is almost trivial once we understand that we can order our set of eigenvalues and combining this with the following notation:
$$v_2 = \text{arg} \max_{v \in \mathbb{R}^p, v \perp v_1 } \frac{v^T X^T X v}{v^T v} $$
All that this is asking, is that we find the eigenvector that maximizes the quotient with respect to all vectors in $\mathbb{R}^p$ which are orthogonal to $v_1$. Furthermore, this corresponds to taking the second largest eignevalue from our ordered set. However, we will show this more rigorously using the argument made in class. We begin with the spectral decomposition of our feature matrix. 
$$X^T X = UDU^T \text{ and define } Z = U^Tv \text{ where } v \in \mathbb{R}^p $$
$$ \Rightarrow v^T X^T X v = z^T D z = \sum_{i=1}^{p} \lambda_i z_i z_i^T \leq \sum_{i=1}^{p} \lambda_2 z_i z_i^T = \lambda_2 \sum_{i=1}^{p} z_i z_I^T = \lambda_2 v^T U^T U v$$
The inequality above turns into an equality in the case where $z_3 = z_4 = \ldots = z_p = 0$; otherwise we can say that $\lambda_2 > \lambda_i$ for $i \in \{3, 4, \ldots, p\}$ which in turn maximizes the quotient with respect to the constraints. This concludes the proof.

### Question 4 
Givens: $\lambda^* = \{14.1, 4.3, 1.2, 0.4 \}$ and $v_1^T = [0.39, 0.42, 0.44, 0.69]$, $v_2^T = [0.40, 0.39, 0.42, -0.72$ 
\newline

#### Part A
We are asked to calculate the proportion of variance explained by the fist 2 principle components.
$$\text{proportion explained} = \frac{\sum_{i=1}^{2} \lambda_i }{\sum_{i=1}^{4} \lambda_i } = \ldots = 0.92$$

#### Part B
Interpreting the principle components: 
1. Noticing that each of the components have a positive coefficient we can say that heavier and longer burritos have a tendency to be tastier. \newline
2. Here we see that the component corresponding to taste has a negative coefficient which indicates that we have some burritos that are small and tasty OR large and bland (their scores would be equivalent)

#### Part C
Here we are tasked with reconstructing the variance-covariance matrix of the centred data based on our 2 eigenpairings. This is a simple process and we back out the desired result from the spectral decomposition of the variance-covariance matrix (yes, this is a symmetric positive definite matrix and we will prove it below).
$$\Sigma = UDU^T = \sum_{i=1}^{2} \lambda_iv_i v_i^T = \lambda_1 v_1 v_1^T + \lambda_2 v_2 v_2^T  $$
We notice that we have all of the quantities specified above so now its a matter of matrix multiplication. I will spare you all of the gory details and skip straight to the answer. 

$$\Sigma = \left( \begin{array}{cccc} 
2.833 \\ 
2.980 & 3.141 \\
3.142 & 3.310 & 3.488 \\
2.556 & 2.879 & 2.980 & 8.942
\end{array} \right) $$
Noting that $\Sigma$ is symmetric we can fill in the rest. As an extension to the problem, we can multiply our eigenvectors by our loading factors matrix to get estimates of mean corrected data.
$$X = LV^T \Rightarrow x_i^T = \left( \begin{array}{cc} l_1 & l_2 \end{array} \right) 
\left( \begin{array}{c} v_1^T \\ v_2^T \end{array} \right) = \ldots = \left( \begin{array}{cccc} 0.39l_1 + 0.40l_2 & \ldots & 0.69l_1 -0.72l_2 \end{array} \right)$$

## Week 5 - A Review 
This week is dedicated to reviewing some of the material we have covered so far (clutering algorithms, linear regression, PCA). These concepts will appear on the midterm and the style of problem is likely to be similar. I am only posting solutions to the proof questions and omitting any short answer/discussion based ones. Furthermore, the WTS lines are abreviated relative to the question on the problem set; needless to say, it encompasses everything that the original question asks for,

### Question 1
#### Part B
\textbf{WTS:} K-means algorithm converges 
\newline
\textit{Proof:} Fix K points from your data set (sample them randomly). We know that we now iterate through the following steps until we reach our "exit condition" of our cluster centres being unchanged with any additional iterations (i.e. we have the most optimal clusters with respect to our cost function). As a recap, here are the steps: 
\newline
\begin{enumerate}
  \item Assign each data point to a cluster based on the a loss function. We define $l(\cdot) = \min_{i\in \{1,\ldots, k\}} ||x-z_i||^2_2$ (the shortest Euclidean distance between a point and one of the K centres.)
  \item Recompute the centre $z_i$ by taking the arithmetic mean of all points that have been assigned to a respective cluster. 
\end{enumerate}

Now we define our cost function. $$cost = \sum_{i=1}^{k} \sum_{x \in C_i} ||x-z_i||^2 = \sum_{i=1}^k \min_{i\in \{1,\ldots, k\}} ||x-z_i||^2$$
We now need to minimize this function with respect to $z_i$. I want to emphasize that previous statement because it dictates exactly how we approach our optimization. We proceed by derivating the cost function.
$$\nabla_{z_i}cost = -2 \sum_{i=1}^{k} \sum_{x \in C_i}(x-z_i) $$
Now, because we are minimizing with respect to an \textit{individual} cluster centre we are not interested in in that outer summation so we can drop it all together. 
$$0 = \sum_{i=1}^{k} \sum_{x \in C_i}(x-z_i) = \sum_{x \in C_i}x - \sum_{x \in C_i} z_i \Rightarrow \sum_{x \in C_i}x = \sum_{x \in C_i} z_i \Rightarrow z_i^* = \frac{1}{|C_i|} \sum_{x \in C_i} x$$
where $|C_i|$ is the number of points within the $i^{th}$ cluster. Notice how this is simply the arithmetic mean of the cluster. We now move on to examine the second derivative. 
$$\nabla^2_{z_i}cost = -2 \sum_{x \in C_i} (-1) = 2|C_i| >0 $$
As expected, this value corresponds to the global minimum of the function. To complete the proof we now notice that we have a function that attains a minimum and is always decreasing. Additionally the $cost \geq 0$ since it is a sum of distances (always positive). Combining these two facts, we get convergence. This completes the proof.

### Question 2
#### Part A
\textbf{WTS:} least variation explained corresponds to smallest eigenvector
\newline
\textit{Proof:} This proof is similar to the most variation explained one, we just take the opposite bound. 
$$X^T X = UDU^T$$ and define $$
Z = U^Tv \text{ , } v \in \mathbb{R}^p $$

$$ \Rightarrow v^T X^T X v = z^T D z = \sum_{i=1}^{p} \lambda_i z_i z_i^T \geq \sum_{i=1}^{p} \lambda_p z_i z_i^T = \lambda_p \sum_{i=1}^{p} z_i z_I^T = \lambda_p v^T U^T U v = \lambda_p v^Tv$$
The equality holds only in the case were $\{ \lambda_1, \ldots, \lambda_{p-1} = 0 \}$, which ensures that this is the minimum and the corresponding eigenvector will explain the least amount of variation.

#### Part B
\textbf{WTS:} proportion of variance explained by jth component is $\frac{\sigma_j^2}{\sum_i \sigma^2_i}$
\newline
\textit{Proof:} This proof is reduced to one line once we notice that 
$$\mathbb{V}(X)= \text{Tr}(\Sigma) = \sum_i \sigma^2_i \Rightarrow \text{prop} = \frac{\sigma_j^2}{\sum_i \sigma^2_i}$$

### Question 4

#### Part A
\textbf{WTS:} $||Y-X\beta||^2$ is minimized at $X^T X \beta =X^T Y$
\newline
\textit{Proof:} We will avoid the gory details of this proof by using matrix calculus. However, before we dive into that I want to take a moment to explain how to approach not only this problem. but proofs in general. 
\newline 
What we must remember is that all of these proofs are fundamental in statistics and as such we can rely on our most basic of concepts in order to arrive to the conclusions. In this case, we know that linear regression written in matrix form is $Y = X\beta + \epsilon$. In simple linear regression, we solved for $\{ \beta_0 , \beta_1 \}$ by minimizing the least squares error. We will try the same approach here (spoiler: it works!). When working with matricies, we know that the equivalent to squaring something is left multiplying by its transpose. With this being said, we can proceed. 
\begin{equation}
  \epsilon^T \epsilon = (Y-X\beta)^T (Y-X\beta) \\
  = (Y^T-\beta^T X^T) (Y-X\beta) \\
  = Y^T Y -\beta^T X^T Y - Y^T X\beta + \beta^T X^T X\beta
\end{equation}

Now we want to optimize this with respect to $\beta$ since this is the parameter we want to calculate from our data (our data are n-tuples of x's and a y). To do so we will use the matrix calculus mentioned above, specifically the following theorems (you are encouraged to pr prove these on your own so you are convinced).

\begin{enumerate}
  \item $Y+AX \Rightarrow \partial_X Y = A$
  \item $\alpha = Y^T A X \Rightarrow 1. \partial_X \alpha = Y^T A \text{ & } 2. \partial_Y \alpha = X^T A^T$
  \item $\alpha= X^T A X \Rightarrow \partial_X \alpha = X^T (A + A^T)$
\end{enumerate}

Here is a [link](https://atmos.washington.edu/~dennis/MatrixCalculus.pdf) to verify your proofs. We can now proceed to differentiate the expansion above and find the critical value.
\begin{equation}
  \nabla_\beta (\epsilon^T \epsilon) = 0 -(X^T Y)^T - Y^TX + 2 X^T X \beta \\
  0 = -2X^T Y + 2X^T X \beta \\
  X^T X \beta = Y^TX
\end{equation}
This completes the proof. Now please note that we can use the above formula to calculate our estimate for $\beta$ given a data set. All we need to do is write it explicitly in the form 
$$\hat{\beta} = (X^T X)^{-1} X Y^T $$

This concludes this week's review, best of luck with the midterm.

## Week 6
This week served as a work period to go over any coding related questions in class. 

## Week 7
Classification problems are those where you have to assign a random to a group. The way we do this, is by examining the features of a new input and use our classification rule which we establish using our data. For more details on this please see the lecture notes. It will also have a mathematical expresion for this process.
\newline 
We now go over the problems in the assignment.
### Question 1

For any classification function $h$, the expected loss can be given directly as:

$$R(h) = \alpha_{FN}\mathbb{P}(h(x)=0,y=1|X=x) + \alpha_{FP}\mathbb{P}(h(x)=1, y = 0 | X=x) \\
= \alpha_{FN}\mathbb{P}(y=1|h(x)=0,X=x)\mathbb{P}(h(x)= 0|X=x) + \\
\alpha_{FP}\mathbb{P}(y=0|h(x)=1,X=x)\mathbb{P}(h(x)=1|X=x)$$
Now, writing $m(x) = \mathbb{P}(y=1 | X=x)$ we can rewrite this as:

$$R(h) = \alpha_{FN}m(x)(1-h(x)) + \alpha_{PN}(1-m(x))h(x)$$

Now let $h^*$ be the optimal classifier. And let $h$ be any classifier.
We know that $R(h^*)-R(h) \leq 0$. And then:

$$R(h^*)- R(h) = -\alpha_{FN}m(x)h^*(x)+\alpha_{FP}h^*(x) - \alpha_{FP}m(x)h^*(x) \\
+ \alpha_{FN}m(x)h(x) - \alpha_{FN}h(x) + \alpha_{FP}m(x)h(x) = \\
= (h^*(x) - h(x))[\alpha_{FP}-(\alpha_{FN}+\alpha_{FP})m(x)]= \\
(\alpha_{FN}+\alpha_{FP})(h^*(x) - h(x))(\frac{\alpha_{FN}}{\alpha_{FN}+\alpha_{FP}} - m(x)) \leq 0$$
We can see that his is exactly $0$ when:

$$\mathbb{P}(y=1 | X=x) = m(x) = \frac{\alpha_{FN}}{\alpha_{FN}+\alpha_{FP}}$$
### Question 2
This question generalizes logistic regression to the case where we have K labels/groups. To do so, we set the Kth class as the baseline and use a K-1 encoding system to solve the problem. We are given the log-odds ratio for the i-th class.

$$ \log \frac{\mathbb{P} (G = i | X = x) }{\mathbb{P} (G = K | X = x)} = \beta_{i,0} + \beta_i^Tx $$
We now exponentiate to get rid of the log. 
$$ \frac{\mathbb{P} (G = i | X = x) }{\mathbb{P} (G = K | X = x)} = \exp \{\beta_{i,0} + \beta_i^Tx \} $$
We can sum the above over all possible values of i.
$$\frac{ \sum_{i=1}^{K-1}\mathbb{P} (G = i | X = x) }{\mathbb{P} (G = K | X = x)} = \sum_{i=1}^{K-1} \exp \{\beta_{i,0} + \beta_i^Tx \} $$
The numerator of the left hand side can we simplified if we apply the law of total conditional probability with respect to G. 
$$ \sum_{i=1}^{K-1}\mathbb{P} (G = i | X = x) = 1 - \mathbb{P} (G = K | X = x) $$
$$ \frac{1}{\mathbb{P} (G = K | X = x)} - 1 =  \sum_{i=1}^{K-1} \exp \{\beta_{i,0} + \beta_i^Tx \} \Rightarrow  \mathbb{P} (G = K | X = x) = \frac{1}{ 1 + \sum_{i=1}^{K-1} \exp \{\beta_{i,0} + \beta_i^Tx \}}$$
From the above, we can apply Bayes' Theorem to get the conditional probabilty of a single class.
$$ \mathbb{P} (G = j | X = x) = \frac{ \exp \{\beta_{j,0} + \beta_j^Tx \}}{ 1 + \sum_{i=1}^{K-1} \exp \{\beta_{i,0} + \beta_i^Tx \}}$$
This completes the proof.

### Question 3
Lab 4.6.2 from the textbook: implementation of logistic regression model.
```{r, eval=TRUE, echo=TRUE, message=FALSE, error=FALSE}
suppressMessages(library(tidyverse))
# load the Smarket data from the ISLR library 
# if you need: install.packages("ISLR")
library(ISLR)
head(Smarket)

# logistic regression is encompassed by a larger class of models
# called Generalized Linear Models. As such we will fit using the build in glm() 
# function and specify that we want the logist subcategory
logit_model <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Smarket, family = binomial)
# using ?glm in your console you can see other types of family's for GLMs but you
# need to first understand link functions 
summary(logit_model)
```

This gives us the model for the data. We can explore this further and clean it up to make it "better". Start by examining the coefficients and the respective P values in order to see if we can drop some of them to get a simpler model.

```{r, eval=TRUE, echo=TRUE, message=FALSE, error=FALSE}
print("The coefficient of each predictor is:")
round(coef(logit_model),4)

print("The summary statistics for each predictor are:")
round(summary(logit_model)$coeff, 4)

```

We immediately see that we cannot reject any of the predictors are an $\alpha = 0.5$ level since the P values are quite high. Let's examine what our model predicts for each of the data points we fed into it based on the parameters we computed.
```{r, eval=TRUE, echo=TRUE, message=FALSE, error=FALSE}
# use the predict function to predict on all the data that was fed into it, 
# or we could specify a subset. This is useful when we have training and testing sets
logit_pred <- predict(logit_model, type = "response")
round(logit_pred[1:10], 4)
# Assuming a hard cutoff rule of 0.5 then we can extrapolate 0's and 1's from this
as.numeric(round(logit_pred[1:10], 4) > 0.5)
```


## Week 8 - Midterm Practice
### Question 1

#### Part A
Loss function for linear regression is $L(\beta,x,y) = (y-x^T\beta)^2$.   
Loss function for classification is $L(h(\cdot),y,x) = I(h(x) \neq y)$.

#### Part B 
We could make the **generalization** that fitting an ML model on just STA314 students if we were certain that it accurately represented the population of the statistics major. However, we know that this isn't the case since not all students in the major are required to take this course. If we train our model on just this subset of the data we would be training the model to pick up many of the specific features of students in STA314. The key here is to realize that our models cannot extrapolate, but only interpolate within the space defined by our training data.

### Question 2
#### Part A
This question asks us to compute the single and complete linkage clusterings. These calculations will be omitted as they are outlined in the lecture slides. Here are the second to last merges for each linkage method to check your answers.  
1. Single linkage - $\{(51, 52.53.101, 103), (102)\}$  
2. Complete linkage = $\{(51, 53.103), (52, 101, 102) \}$  

#### Part B
K-means++ attempts to remediate the situation in which the random sampling that sets the first centres has them too close together by taking K points that are as far appart from each other as possible. But note that this isn't very efficient since you need to compute and compare distances between every point of your data set.

### Question 3

#### Part A
Here are the conditions for implementing linear regression.
1. Linearity - Plot the residuals and look for nonlinearities  
2. Independence of errors - see tests for indepedence from STA261 and STA302  
3. Identically distributed errors - plot errors and check for homoskedasticity  
4. Normality of errors - QQ-plot and empirical CDF assessment

#### Part B
Let $A = uv^T$. Then the eigenvalue equation $Ax= \lambda x$ becomes
\begin{align}
Ax &= \lambda x \notag\\
uv^Tx &= \lambda x \notag\\
(v^Tx)u = \lambda x.\label{eqn1}
\end{align}
Now, $v^Tx$ and $\lambda$ are both scalars, so the only way the final vector equality can hold is if $x = \alpha u$, for some $\alpha \in \mathbb{R}$.  By convention, eigenvectors have unit norm, so the eigenvector is 
$$x = \frac{u}{\|u\|}$$

Substituting that into the above, we get 
$$\frac{v^Tu}{\|u\|} u = \frac{\lambda}{\|u\|} u,$$ 
which implies $\lambda = v^Tu$.  To show that this is the largest eigenvector (in absolute value), we note that if we choose $x$ to be any of the $n-1$ vectors $w$ that satisfy $v^Tw=0$, then $Ax=\lambda x$ simplifies to $0\cdot w = \lambda w$, which implies that all of the other eigenvalues of $A$ are zero.



### Question 4
#### Part A
We want to find the direction that explains the **third** largest amount of information.  Given that we already know the first two PCs, we don't need to explain any more information in those particular directions, so we're only interested in directions that are **orthogonal** to $v_1$ and $v_2$.  From the lectures, we know that the variance in a direction $u$ is given by 
$$(n-1)^{-1}\|Xu\|$$
So we need to solve the following optimization problem 
$$v_24= \arg \max_{\substack{u^Tv_1=0\\ u^Tv_2 =0 \\ \|u\|=1}} u^TX^TXu = \arg \max_{\substack{u^Tv_1=0\\ u^Tv_2 =0 }} \frac{u^TX^TXu}{u^Tu} $$

Following the same reasoning as the lectures, we compute the eigendecomposition $X^TX = V\Lambda V^T$ and set $z = V^Tu$. Because $u \perp v_1, v_2$, the first two components of $z$ will be zero.  Then it follows that 

$$ u^TX^TXu = z^T\Lambda z = \sum_{i=1}^p \lambda_iz_i^2 =\sum_{i=3}^p \lambda_iz_i^2 \leq \lambda_3\sum_{i=3}^p z_i^2  = \lambda_3 u^TVV^Tu = \lambda_3 u^Tu $$

As in class, that inequality is an equality if $z_4= z_5 = \ldots = z_p=0$, which occurs when $u = v_3$. As we have shown that the Rayleigh quotient cannot be larger than $\lambda_3$ when computed on vectors that are orthogonal to $v_1$ and $v_2$ **and** that value is attained when $u=v_3$, it follows that $v_3$ maximizes the constrained Rayleigh quotient and hence the second principal component is the eigenvector of $X^TX$ that corresponds to the second largest eigenvalue. This completes the proof.

#### Part B
In forward variable selection you start with a model that only has an inter- cept. At each step you add the feature that maximizes the prediction error on the test set. Continue until all features have been added.  
In backward variable selection, you start with a model that has all the features in it. At each step you remove the feature that reduces the prediction error the least on the test set. Continue until there are no features left.  
Forwards selection is preferable if there are more features than observations. Best of luck on the midterm.
