---
title: "MTH380 R Sample Code"
author: "Alin Morariu"
date: '2019-09-09'
output: pdf_document
toc: TRUE 
number_sections: TRUE 
fig_caption: TRUE
df_print: tibble 
fontsize: 12pt 
header-includes:
   - \usepackage{amsmath}          
---
# Week 1

## Welcome to MTH380! 

First of, my apologies for not having a projector ready for this week's class. It will not be an issue going forward. This tutorial is dedicated to introducing you to R, a statistical programing language, along side the R Studio IDE. **NOTE: this course does NOT have a programing component to its evaluation. Any material covering R is exclusively for your knowledeg.** Please do not be scared or drop the course based on this tutorial. 

With that out of the way, let's begin. 

### Installing R

The first step is to install a version of R. The simplest way to go about this is to Google "R" and follow the links. I suggest you download the [CRAN](FILL IN LATER) as it saves you some additional set up headaches. Follow this up by installing the *free* version [R Studio](FILL IN LATER). R Studio acts a "wrapper" for R which makes the user's life much simpler by avoiding having to run scripts through your terminal. If you do prefer executing scripts in the terminal, then feel free to do so but for most its not a natural method of executing code. 

### What you see when you open R Studio

Once you open up R Studio you will see 4 panes (in no particular order):

1. **Console** - think of this as your playground; you can experiment with code and try things out without causing problems or introducting error into your main file
2. **.Rmd or .R File** - speaking of the main file, this is where your "real" work will happen. You will perform your computations, fit models, evaluate data sets, visualize data, and much more in this file. I have provided a [template](INSERT LINK) for a rendering a PDF document (word and HTML documents are also supported, you can change this in the YAML header at the top of the document). 
3. **Environment** - this is where you can see your global memory and what you have stored. Once you load in a dataset, it will show up here and you can start working with it.
4. **File, Plots, Etc.** - this last pane is used to browse your computer's file structure, preview your plots, manage packages, amongst it's other functionalities (not all of which will be useful to you). 

The first step to writing some R code is to create an R script or Rmarkdown file. I suggest going with an Rmarkdown file since you can't output R scripts as a PDF. Here is the [source code](INSERT LINK) for this week's tutorial. Once you have a new file created, we can insert of a chunk of code from the *Code* drop down menu. Alternatively you can use the keyboard shortcuts (Cmd-Alt-I for Mac users, Ctrl-Alt-I for Windows). Inside these chunks will be executable R code which will be run and outputed. 

### Some basics

Here is our first code of chunk going over some of the basic objects available for storing information in R. 

```{r This is a code chunk, eval=TRUE, echo=TRUE, message=FALSE, error=FALSE}
# creating R objects 
# assign something a value with the <- operator, set parameter values with =
x <- 42 # an integer
class(x)

x1 <- "42" # a character
class(x1)

x2 <- c(1, 42, 3, 59, 75, 35, 4, 42) # a vector
class(x2)
x2[3:5] # subsetting vectors (this is very useful)

x3 <- list(x, x1, x2) # a list
class(x3)
x3[[2]] # double brackets to access list elements
x3[[3]][2:5] # accessing elements of a list element 
```

Once I leave my code chunk, I can go back to typing text to comment on the results of my code chunks. You can also include TeX when typing which makes this very nice for writing up assignments and paper (note that it is limited).

```{r Computation with R, eval=TRUE, echo=TRUE, message=FALSE, error=FALSE}
3+2
3-1
3*2
3/2
3^2
log(3) # you can find more as you need

# Matrix operations 

c(1,2,3) %*% c(4,5,6) # standard matrix product (i.e. dot product)
class(c(1,2,3) %*% c(4,5,6))

c(1,2,3) %*% t(c(4,5,6)) # "cross" product since we have the transpose 
class(c(1,2,3) %*% t(c(4,5,6)))
```

### R Packages

One of the nice things about R is that it is an open-source language meaning that anyone can contribute to it. This means that if there is something you'd like to do in R, someone has likely written code to do that for you already and all you have to do is pull in that code to use it. Useful code is often wrapped up as a package which we can install in the following way.

```{r Installing Packages, eval=FALSE}
install.packages("Matrix") # only need to install the package once

library(Matrix) # to use the package you have to call it via the library() function
# you must call it once in ANY file you want to use the functions in the package 
```

```{r}
# using the functions 
crossprod(c(1,2,3), c(4,5,6)) # R knows to search all functions
Matrix::crossprod(c(1,2,3), c(4,5,6)) # can specify which package to get a function from
# this second method is great to produce more explicit code
```

### Extracting information from data

For this section, we will be working with the built in Iris data set created by Ronald Fisher and Edgar Anderson in the 1930s. 

```{r Iris analysis}
# preview the data set
head(iris, 10) # option 1 and my favourite
dplyr::glimpse(iris) # option 2, works better than head on larger data sets

summary(iris) # summary statistics of Iris dataset
summary(iris$Petal.Length) # summary statistics of a subset of Iris data set

petal.summary <- summary(iris$Petal.Length)
class(petal.summary) # a vector! so we can access specific elements

length(petal.summary)
petal.summary[1]
petal.summary[2] # and so on unitl...
petal.summary[6]
petal.summary[7] # this doesn't exist! 
```

### Plotting our data 

This section will focus on plotting the data from the Iris dataset with tidyverse. As a part of the tidyverse, there is a smaller packaged called *ggplot2* (stands for grammar of graphic) which is an incredibly powerful tool for data visualization. It is far superior to the base R plot function since it works off of tibbles (a tidyverse specific data structure) which allows for ggplot to build plots by layering. This gives the user much more control over what is and isn't shown and makes for more generic code that can be used across many projects up to variable names.

```{r ggplot introduction}
library(ggplot2) 
library(cowplot)

iris <- as_tibble(iris)
head(iris)

# let's make a histogram
iris.hist <- ggplot(iris, aes(Sepal.Length)) + 
     geom_histogram(bins = 10, color = "navy", fill = "steelblue") + 
     labs(title = "Histogram of Sepal Lengths",
          subtitle = "Data from R. Fisher and E. Anderson", 
          xlab = "Sepal Lenght (cm)",
          ylab = "Counts") + 
     theme_minimal()
iris.hist

# let's make a bar graph (think about the difference between a bar graph and historgram)
iris.bar <- iris %>% ggplot(aes(Species)) +
     geom_bar(color = "navy", fill = "steelblue") + 
     labs(title = "Bar graph of Iris Species",
          subtitle = "Data from R. Fisher and E. Anderson", 
          xlab = "Species",
          ylab = "Counts") + 
     theme_minimal()
iris.bar 

# let's make a dot graph with a line of best fit
iris.line <- iris %>% ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
     geom_point(aes(color = factor(Species)), color = "steelblue") + 
     geom_smooth(method='lm', color = "gold") + 
     labs(title = "Sepal Width vs. Sepal Length",
          subtitle = "Data from R. Fisher and E. Anderson", 
          xlab = "Sepal Length",
          ylab = "Sepal Width") + 
     theme_minimal()
iris.line

# since we stored each graph in memory we can arrange them nicely
cowplot::plot_grid(iris.hist, iris.bar, iris.line, 
     ncol =2, 
     labels=LETTERS[1:3])
```

For more examples on graphing with ggplot2, check out the [ggplot flipbook](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#1) by Gina Reynolds.

### Working with probability distributions

For the purpose of this course, the most useful component of R is working with probability distributions. We will formally define probability distributions and provide examples throughout the duration of the course but as an example for today I will focus on the normal/Gaussian distribution which takes the following form

\begin{align}
     f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp{ \left( - \frac{(x - \mu)^2}{2 \sigma^2} \right) }
\end{align}

Where $\mu$ is the mean of the distribution and $\sigma$ is the standard deviation. A multivariate form exists but will be left out for now (more on multivariate distributions in later tutorials). So now for how R can help us make sense of these distributions. We can evaluate them at various points and compute probabilities of certain events happening. Or we can generate random samples from a the distribution for simulation purposes. 

```{r Working with distributions}
# generate random normals 
round(rnorm(n = 25, mean = 4, sd = 1.2), 2)

# compute probabilities 
dnorm(4.67, mean = 4, sd = 1.2) # value of the density (i.e. plug in x = 4.67 into f(x))
pnorm(4.67, mean = 4, sd = 1.2) # probability of x = 4.67 (i.e. the integral over f from - infinity to 4.67)
```

This should give you a solid idea regarding what you can do with R. Throughout the semester, I will try to highlight various theorems which you could verify empirically using these tools. Additionally, there will be more sample code added to this page for you to try out on your own and learn from. 
